---
title: "Note"
output: html_document
---

```{r}
library(tidyverse)
```


# cours 1


Ctrl + Space (auto-completion, better than Tab)
Ctrl + Up (command history & search)
Ctrl + Click (function source code)
Ctrl + Enter (execute line of code)
Ctrl + Shift + A (reformat code)
Ctrl + Shift + C (comment/uncomment selected lines)
Ctrl + Shift + K (knit)
Ctrl + Shift + B (build package, website or book)
Ctrl + Shift + M (pipe)
Alt + Shift + K to see all shortcuts…

# cours 2

les methods sont des fonctions pour certaine classe

apres un lapply faire un do.call

```{r}
# lapply et do.call
tmp1 <- lapply(1:10, function(x) rnorm(10, x))
tmp2 <- do.call("rbind", tmp1)

# replicate
a <- replicate(5, rnorm(10))
```

order peut ordonner un data.frame 

```{r}
order(c(1, 6, 8, 2, 2), c(0, 0, 0, 2, 1))
# et different de 
order(c(1, 6, 8, 2, 2, 0, 0, 0, 2, 1))
```

Pour ordonner des caracteres selon les nombres --> interressant
gtools::mixedsort(c("a1", "a2", "a10"))  ## not in base, obviously

```{r}
gtools::mixedsort(c("a1", "a2", "a10"))
```

pmin et pmax --> min et max vectorisé

outer --> 
```{r}
outer(1:4, 1:3,'+') # la fonction entre '' doit pouvoir etre vectorisé
```

```{r}
expand.grid(param1 = c(5, 50), param2 = c(1, 3, 10))
```

```{r}
me <- 1:10
a <- glue::glue("I am {me}")  ## not in base, but SO useful
```

pour lister une liste de fichier
```{r}
x <- list.files(pattern = "\\.Rmd$", full.names = TRUE)
```

intersect, union et diff

```{r}
intersect(letters[1:4], letters[3:5])
union(letters[1:4], letters[3:5])
setdiff(letters[1:4], letters[3:5])
```

```{r}
# EXO 3.4.5
# 1

x <- 1:100
k <- 3

y <- sample(rep_len(1:k, length(x)))
ls <- split(x, y)
# 2
set.seed(1)
x <- rnorm(10)
mean(x)
error <- qt(0.95, df = length(x) - 1 ) * sd(x) / sqrt(length(x))
c(mean(x) - error, mean(x) + error)

x <- replicate(100, sample(x, length(10), replace = T))

# pour le bootstrap
boot <- replicate(10000, mean(sample(x, replace = T)))
quantile(boot, probs = c(0.25, 0.975))


# 3
my_mtcars <- mtcars[c("mpg", "hp")]
my_mtcars$my_col <- sample(c("mpg", "hp"), size = nrow(my_mtcars), replace = TRUE)
head(my_mtcars)

a <- cbind(1:nrow(my_mtcars), match(my_mtcars$my_col, names(my_mtcars)))
cbind(my_mtcars, truc = my_mtcars[a])


# 4

df <- data.frame(
  id1 = c("a", "f", "a"),
  id2 = c("b", "e", "e"), 
  id3 = c("c", "d", "f"),
  inter = c(7.343, 2.454, 3.234),
  stringsAsFactors = FALSE
)
df

code <- setNames(1:6, letters[1:6])

df[1:3] <- lapply(df[-4], function(var) code[var])


```

les class

```{r}
agent007 <- list(first = "James", last = "Bond")

class(agent007) <- "Person"  ## "agent007" is now an object of class "Person"
# Just make a function called <method_name>.<class_name>()
print.Person <- function(x) {
  print(glue::glue("My name is {x$last}, {x$first} {x$last}."))
  invisible(x)
}

agent007


Person <- function(first, last) {
  structure(list(first = first, last = last), class = "Person")
}
me <- Person("Florian", "Prive")

Worker <- function(first, last, job) {
  obj <- Person(first, last)
  obj$job <- job
  class(obj) <- c("Worker", class(obj))
  obj
}
print.Worker <- function(x) {
  print.Person(x) 
  print(glue::glue("I am a {x$job}."))
  invisible(x)
}

(worker_007 <- Worker("James", "Bond", "secret agent"))
```

# Cours 3

--> les tiddle --> sorte de data.frame

```{r}
library(tidyverse)

as_tibble(iris)

tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)

tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)

# exo

df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

ti <- tibble(abc = 1, xyz = "a")

ti$xyz
ti[c("xyz")]
ti[c("abc", "xyz")]

a <- as_tibble(mtcars)
var <- "mpg"

a[var]


annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

a1 <- annoying$`1`

plot(annoying)

annoying$`3` <- annoying$`2` / annoying$`1`

colnames(annoying) <- c("one", "two", "three")

tibble::enframe(annoying)


# aes_string pour faire des boucle en ggplot --> permet de prendre des var en X
# cowplot::plot_grid() a la place de grid_extra
# cowplot:: get_legend()

```

```{r}
(df <- gapminder::gapminder %>%
   filter(year == 1992))

# gdp en fonction de lifeEXP


ggplot(df, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10() +
  xlab("Gross Domestic Product (log scale)") +
  ylab("Lite Expectancy at birth (years)") +
  ggtitle("Gapminder for 1992") +
  labs(size = "Population", color = "Continent") +
  theme_bw() +
  theme(axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5), 
        title = element_text(size = 15, face = "bold"))



ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )


as_tibble(diamonds)

ggplot(diamonds, aes(cut, depth)) +
  geom_pointrange(stat = "summary")

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 0)

x <- rnorm(100000)
y <- 2 * x + rnorm(100000, sd = 2)
df <- data.frame(x = x, y = y)

ggplot(data = df, mapping = aes(x, y)) + 
  geom_point(alpha = 0.1)

```

# Cours 4

```{r}
library(nycflights13)
library(tidyverse)

flights

jan1 <- filter(flights, month == 1, day == 1)

# fonction near() pour tester une egalité sans probleme de nombre flottant

nov_dec <- filter(flights, month == 11 | month == 12)
# equivalent a :
nov_dec <- filter(flights, month %in% c(11, 12))

# les NAs
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)

filter(df, is.na(x) | x > 1)

# exo

(flights)

ti <- filter(flights, arr_delay >= 2 * 60) 
ti <- filter(ti, dest %in% c("IAH", "HOU"))

# AA      American Airlines Inc.
# DL      Delta Air Lines Inc. 
# UA      United Air Lines Inc.    

ti <- filter(ti, carrier %in% c("AA", "DL", "UA"))
ti <- filter(ti, month %in% c(7, 8, 9))
ti <- filter(ti, arr_delay >= 2 * 60 & dep_delay == 0)

summary(flights)

# le pipe de dplyr
ti <- filter(flights, arr_delay >= 2 * 60) %>%
  filter(., dest %in% c("IAH", "HOU"))

# on peut faire des calculs
filter(flights, dep_delay >= 60, dep_delay - arr_delay > 30)

arrange(flights, year, month, day)
# permet d'ordonner selon l'annee, le mois et ensuite le jour

# pour ordonner de facon decroissante
arrange(flights, desc(dep_delay))


arrange(flights, desc(is.na(dep_time)))

arrange(flights, desc(arr_delay))
arrange(flights, air_time)

# Select all columns between year and day (inclusive)
select(flights, year:day)
# Select columns by name
select(flights, year, month, day)

# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
select(flights, 1:3)



# selection de lignes
flights %>% 
  slice(10:22)


rename(flights, annee = year)

select(flights, time_hour, air_time, everything())
```

There are a number of helper functions you can use within select():

    starts_with("abc"): matches names that begin with “abc”.

    ends_with("xyz"): matches names that end with “xyz”.

    contains("ijk"): matches names that contain “ijk”.

    matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.

    num_range("x", 1:3): matches x1, x2 and x3.
    
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, vars)

select(flights, contains("TIME"))

flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)

mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)

transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)


transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)


transmute(flights, time = dep_time %/% 100 * 60 + dep_time %% 100)


# truc utile
# le mutate if
mutate_if(iris, as.factor, as.character)

mutate_at(flights, c("dep_time", "sched_dep_time"), function(x) {
  (x %/% 100 * 60 + x %% 100) %% 1440
})
```

# cours 5

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))

by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))

# or

flights %>% 
  group_by(year, month, day) %>%
  summarise(delay = mean(dep_delay, na.rm = TRUE))


by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(), # fonction pour compter le nombre d'occurence
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")


ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")



# Convert to a tibble so it prints nicely
batting <- as_tibble(Lahman::Batting)

batters <- batting %>% 
  group_by(playerID) %>% 
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>% 
  filter(ab > 100) %>% 
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() + 
    geom_smooth(se = FALSE)


not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))


not_cancelled %>% 
  count(tailnum, wt = distance)


daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))



daily %>% 
  ungroup() %>%             # no longer grouped by date
  summarise(flights = n())  # all flights


# exo 
# A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.


# A flight is always 10 minutes late.
flights %>% 
  group_by(flight) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  filter(delay >= 10)


not_cancelled %>% 
  count(dest)

not_cancelled %>% 
  group_by(dest) %>% 
  summarise(n())

not_cancelled %>% 
  count(tailnum, wt = distance)


# Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

cancelled <- flights %>% 
  filter(is.na(dep_delay), is.na(arr_delay))

cancelled %>% 
  group_by(month, day) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(day, n)) +
  geom_point() +
  geom_line() +
  coord_polar()



flights %>% 
  group_by(carrier) %>% 
  summarise(arr_delay = mean(arr_delay, na.rm = T),
            n = n()) %>% 
  arrange(desc(arr_delay)) %>% 
  left_join(airlines) %>% 
  filter(n > 100)
  

popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)


popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
  select(year:day, dest, arr_delay, prop_delay)

flights %>% 
  group_by(tailnum) %>% 
  summarise(worst = max(arr_delay, na.rm = T)) %>% 
  arrange(desc(worst))

# What time of day should you fly if you want to avoid delays as much as possible?
flights

flights %>% 
  group_by(hour) %>% 
  summarise(arr_delay = mean(arr_delay, na.rm = T), n = n()) %>% 
  ggplot(aes(hour, arr_delay, color = n)) +
  geom_point() +
  scale_color_viridis_c()

# For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.

flights %>% 
  filter(arr_delay > 0) %>% 
  group_by(dest) %>% 
  mutate(sum_delay = sum(arr_delay, na.rm = T), prop_delay = arr_delay / sum_delay)

# Database
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")

dbListTables(con)
dbWriteTable(con, "mtcars", mtcars)
dbListTables(con)

dbListFields(con, "mtcars")
dbReadTable(con, "mtcars")

# You can fetch all results:
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(res)
dbClearResult(res)

# Or a chunk at a time
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
while(!dbHasCompleted(res)){
  chunk <- dbFetch(res, n = 5)
  print(nrow(chunk))
}
dbClearResult(res)

DBI::dbDisconnect(con) 

library(dbplyr)
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
copy_to(con, nycflights13::flights, "flights",
  temporary = FALSE, 
  indexes = list(
    c("year", "month", "day"), 
    "carrier", 
    "tailnum",
    "dest"
  )
)
flights_db <- tbl(con, "flights")

flights_db %>% filter(dep_delay > 240)

tailnum_delay_db <- flights_db %>% 
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay),
    n = n()
  ) %>% 
  arrange(desc(delay)) %>%
  filter(n > 100)

tailnum_delay <- tailnum_delay_db %>% collect()

```

# cours 6

--> chapitre important


```{r}
library(nycflights13)
library(dplyr)

# il faut mieux travailler sur les colonnes que sur les lignes (pour un data.frame)
# plus rapide au niveau de la memoire

# ca c'est lent --> temps N au carré
x <- rnorm(1e5)  # Try also with n = 1e5
system.time({
  current_sum <- 0
  res <- c()
  for (x_i in x) {
    current_sum <- current_sum + x_i
    res <- c(res, current_sum)
  }
})

# ca c'est rapide (si on connait la taille du resultat) --> temps N 
system.time({
  current_sum <- 0
  res2 <- double(length(x))
  for (i in seq_along(x)) {
    current_sum <- current_sum + x[i]
    res2[i] <- current_sum
  }
})
# si on connait pas la taille : rapide aussi --> temps N * K
system.time({
  current_sum <- 0
  res3 <- list()
  for (i in seq_along(x)) {
    current_sum <- current_sum + x[i]
    res3[[i]] <- current_sum
  }
})

res3 <- unlist(res3) # pour defaire la liste --> temps N * logN

# ou alors 
system.time({
  current_sum <- 0
  res4 <- c()
  for (i in seq_along(x)) {
    current_sum <- current_sum + x[i]
    res4[i] <- current_sum
  }
})

# ou parfait (car coder en C)
res5 <- cumsum(x)


# autre exemple
n <- 1e3
max <- 1:1000
system.time({
  mat <- NULL
  for (m in max) {
    mat <- cbind(mat, runif(n, max = m))
  }
})

# bon avec une list
system.time({
  l <- vector("list", length(max)) # pour initialiser la liste
  for (i in seq_along(max)) {
    l[[i]] <- runif(n, max = max[i])
  }
  mat2 <- do.call("cbind", l) # penser a l'usage de do.call()
})
# ou un peu pareil
system.time({
  l <- list()
  for (i in seq_along(max)) {
    l[[i]] <- runif(n, max = max[i])
  }
  mat3 <- do.call("cbind", l)
})
# ou : --> pas mal ca
system.time({
  mat4 <- matrix(0, n, length(max))
  for (i in seq_along(max)) {
    mat4[, i] <- runif(n, max = max[i])
  }
})
# ou avec sapply
system.time(
  mat5 <- sapply(max, function(m) runif(n, max = m))
)

# parcourir les matrices en colonnes pour aller plus vite


# package matrixStats --> pour avoir des fonctions style rowMeans() !
# ca va beaucoup plus vite

# lire les .csv --> package data.table::fread(data.table::FALSE)
# fini read.csv


# on peur voir quelle fonction prend du temps
# faire Profile --> Profil Selected Lines ou alt + maj + cmd + P


# Vectorasarion
monte_carlo <- function(N) {
  
  hits <- 0
  for (i in seq_len(N)) {
    u1 <- runif(1)
    u2 <- runif(1)
    if (u1 ^ 2 > u2) {
      hits <- hits + 1
    }
  }
  
  hits / N
}
monte_carlo(1000000)
# traduire la boucle part expression vectorisé

monte_carlo2 <- function(N) {
  
  hits <- 0
  for (i in seq_len(N)) {
    u <- runif(2)
    if (u[1] ^ 2 > u[2]) {
      hits <- hits + 1
    }
  }
  
  hits / N
}

monte_carlo3 <- function(N) {
  mean(runif(N) ^ 2 > runif(N))
}

system.time(a1 <- monte_carlo(1000000))
system.time(a2 <- monte_carlo2(1000000))
system.time(a3 <- monte_carlo3(1000000))


# les boucles en C ++
library(Rcpp)



is_odd_r <- function(n = 10) {
  n %% 2 == 1
}

is_odd_r(1:10)


gibbs_r <- function(N, thin) {
  mat <- matrix(nrow = 2, ncol = N)
  x <- y <- 0
  for (i in 1:N) {
    for (j in 1:thin) {
      x <- rgamma(1, 3, y * y + 4)
      y <- rnorm(1, 1 / (x + 1), 1 / sqrt(2 * (x + 1)))
    }
    mat[, i] <- c(x, y)
  }
  mat
}


gr <- gibbs_r(10000, 100)
plot(t(gr))
```

```{r}
# cours 7
library(tidyverse)

mydf <- readRDS("one-million.rds")

QRA_3Dmatrix <- array(0, dim = c(max(mydf$ID), max(mydf$Volume), 2))

res <- mydf %>%
  group_by(ID, Volume) %>%
  summarise(count = n(), 
            sum_energy = sum(1 - 1.358 / (1 + exp( (1000 * Energy - 129000) / 120300))))

ind <- cbind(res$ID, res$Volume)

QRA_3Dmatrix2 <- array(0, dim = c(max(mydf$ID), max(mydf$Volume), 2))

QRA_3Dmatrix2[cbind(ind, 1)] <- res$count
QRA_3Dmatrix2[cbind(ind, 2)] <- res$sum_energy

# fonction sweep interressante

m0 <- matrix(rnorm(1e6), 1e3, 1e3)

# vectorisation

M <- 50
step1 <- runif(M)
A <- rnorm(M)
N <- 1e4

tau <- matrix(0, N + 1, M)
tau[1, ] <- A
for (j in 1:M) {
  for (i in 2:nrow(tau)) {
    tau[i, j] <- tau[i - 1, j] + step1[j] * 1.0025^(i - 2)
  }
} 

dim(tau)

# du coup 
# matrice M * N (outer multiplication entre 2 vecteurs et tte leurs possibilité)
MN <- rbind(rep(0, M),
            outer(1.0025^(0:(N - 1)), step1))

tau2 <- matrix(A, N + 1, M, byrow = T)

tau2 <- tau2 + matrixStats::colCumsums(MN)

all.equal(tau, tau2)

# exemple outer
a <- (1:10)
b <- (11:1)

ab <- outer(a, b, "-")

x <- 1:9; names(x) <- x
# Multiplication & Power Tables
x %o% x
y <- 2:8; names(y) <- paste(y,":", sep = "")
outer(y, x, "^")

outer(month.abb, 1999:2003, FUN = "paste")

##
x <- sample(10, size = 1e4, replace = TRUE)
breaks <- c(1, 3, 9, 9.5, 10)


hist(x, breaks, plot = F)$count
hist(x, breaks)


n <- length(breaks)
br <- cbind(breaks[1:(n - 1)], 
            breaks[2:n])

a <- rep(NA, nrow(br))
for (i in 1:nrow(br)) {
  a[i] <- sum(x > br[i, 1] & x <= br[i, 2])
}

table(cut(x, breaks))
```

